---
title: "Can't I *please* just visit one friend?"
subtitle: "Visualizing social distancing networks in the era of COVID-19"
author: 
 - Steven M. Goodreau, University of Washington
 - Martina Morris, University of Washington
 - on behalf of the Statnet Development Team
date: "3/31/2020"
output:
  html_document:
    highlight: kate
    theme: lumen
    toc: yes
    toc_float: true
    number_sections: true
    fig_width: 7
    fig_height: 4
    fig_caption: true
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The situation

COVID-19 is raging around the world, and you and your family have been asked (OK, told) to stay home and practice social distancing.

For the most people you are doing so, because you are good people.

But, you wonder - or perhaps more likely your kids wonder - **is there really so much harm in meeting up with just one friend?** 

After all, there are all kinds of interactions still happening - between health care workers and patients, people in the food services industries and their customers, and so on.  Why does one more matter?

This is a question we've heard in a variety of forms, from our own friends and relatives.

For many of the people asking the question, it seems particularly low-risk to meet up with friends if neither person has an elderly member in their household, given how concentrated the mortality is within that population.

As network epidemiologists, this is our effort to explain why these connections matter more than they may seem.

# The Good Old Days

Let's start by imagining back to the good old days, just a few short weeks or months ago, before COVID-19 took off.  People had lots of daily interations with lots of other people They obviously varied in how many they had -- some people had many, and some had few -- but overall the number was a lot higher. 

Now try to imagine what that might look like if you drew it. For network epidemiologists like us, we visualize networks like that using diagrams with *nodes* and *links*.  Here we'll make one in which the nodes represent individual households, and the links represent frequent, repeated interactions between any members of two households. 

Let's say, for the sake of the example, that the average household had 15 interactions with other households. What would that look like?  We can visualize it as a graph of nodes and links, where we draw the nodes (households) as green circles, and the links (interactions between two households) as gray lines. 

Then we get something like this:

```{r pre-COVID, cache=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
library(statnet)
n <- 200
emptynet <- network.initialize(n, directed=FALSE)
meandeg.precov <- 15
fit.precov <- ergm(emptynet~edges, target.stats=meandeg.precov*n/2)
net.precov <- simulate(fit.precov)
plot(net.precov, vertex.cex=1.5, mode="kamadakawai", vertex.col=3, edge.col="gray20")
```

The connections among the nodes are so numerous that you can't even make sense of them - it's all just a single dense mass of ties. For a new virus that is able to spread across those ties, and to which everyone is susceptible, life is pretty good.

# Perfect isolation

Now let's imagine a world at the other extreme - complete and utter lock-down, where somehow we manage to keep every person in their household and never interact with anyone else.  That would look like this:

```{r emptynet, cache=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
plot(emptynet, vertex.cex=1.5, mode="kamadakawai", vertex.col=3, edge.col="gray20")
```

Here the virus would not be able to spread from household to household at all, because there are no links. Anyone who had it might pass it onto the other members of their household, and each of those people would either die or recover.  But the virus could not spread to other households; it would quickly disappear.

# Adding in the essential workers

Of course, that world is impossible. 

Why? Because certain interactions are essential to maintain the safety, health and well-being of all of us:

- We need doctors and nurses and EMTs interacting with people.
- We need all the people that get food from the ground and into our hands (growing it, transporting it, selling it) to interact with people.
- We need the people who manage and maintain our water supply and electricity and gas to keep showing up at work.
- We need first responders keeping us safe.
- In big cities, we need transport workers helping all of these people get back and forth between work and home.
- And so on. 

(Lots of other people have important jobs too, of course, but many of them are able to work from home these days). 

We can aim to minimize the number of direct interactions these people have when they're all doing their thing, and have them follow all of the rules around washing hands or wearing masks if appropriate.  But we can't completely eliminate all risk of transmission here.

So, let's imagine what this might do to our network. It's hard to put real numbers on this, but we'll just create what seems like a reasoanble example. 

Let's say that 1 in every 10 households contains someone with a job where they need to continue some connections. and the average number of connections they have is 4. In reality, they may interact with many more people, but if they are able to use standard advice about hand washing and some distancing throughout the course of their work, let's say they're able to bring their effective contacts down to 4.  That gives us something that looks like this:

```{r essl.net, cache=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
meandeg.essl <- 4
prop.essl <- 0.1
essl <- rep(0, n)
essl[sample(1:n, round(prop.essl*n,0), replace=FALSE)] <- 1
set.vertex.attribute(emptynet, 'essl', essl)
fit.essl <- ergm(emptynet~edges+ nodematch("essl", diff=TRUE, levels=1),
              target.stats= c(
                meandeg.essl*prop.essl*n, 0),
              control=control.ergm(MCMC.burnin = 1e6))
net.essl <- simulate(fit.essl, control = control.simulate(MCMC.burnin = 1e6))
net.essl.el <- as.edgelist(net.essl)
plot(net.essl, vertex.cex=1.5, vertex.col=3+essl, edge.col="gray20")

```

You can see that the virus has some room to move around again.  Not nearly as much as in the days before social distancing, but still some.

Let us be very clear here: some people are going to get infected, and some people are going to die, because of these connections.  It's that simple.  

But these connections are so essential that we as a society are willing to make that trade-off. Without them, many more people would die of other basic things besides COVID-19: starvation, cold, crime, other diseases.

And hopefully, the deaths here won't be that high.  Most households are still effectively isolated, and those that are connected are only held together by a pretty sparse set of ties. Since transmission is not 100% guaranteed, even when two people are in close contact, that means that the sparse network still gives a lot of protection to many of the people in it.

# Visting just one friend

But, we all know now: strict social distancing starts to get boring after a while. And meeting up to hang out with just one person outside your household - a friend especially - is pretty tempting.  And it just doesn't seem like such a big deal when these other connections are already happening, right?

Remember, though, that most households have more than one person in them. In fact, the average household size in the US is 2.6 people.

Let's say, then, that you decide to meet up with your friend Sue.  But if Sue's parents are letting her go out, then they probably also need to let her brother Ed go out.

Let's say, then, that an average of two people in each household decide to maintain connections with one person.  This is an average- some households will have none, and some will have 3, but on average it's 2. Note that this is still less than the average household size - that means that not everybody is meeting up with a friend.

So let's visualize what would happen under that scenario:


```{r non.essl.net.1, cache=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
meandeg.non.essl.1 <- 2
fit.non.essl.1 <- ergm(emptynet~edges, target.stats= meandeg.non.essl.1*n/2)
net.non.essl.1 <- simulate(fit.non.essl.1)
net.comb.1 <- net.non.essl.1
net.comb.1 <- add.edges(net.comb.1, tail=net.essl.el[,1], head=net.essl.el[,2])
plot(net.comb.1, vertex.cex=1.5, vertex.col=3+essl, edge.col="gray20")
```


# Visting just one friend per household

```{r non.essl.net.2, cache=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
meandeg.non.essl.2 <- 1
fit.non.essl.2 <- ergm(emptynet~edges, target.stats= meandeg.non.essl.2*n/2)
net.non.essl.2 <- simulate(fit.non.essl.2)
net.comb.2 <- net.non.essl.2
net.comb.2 <- add.edges(net.comb.2, tail=net.essl.el[,1], head=net.essl.el[,2])
plot(net.comb.2, vertex.cex=1.5, vertex.col=3+essl, edge.col="gray20")
```

# Reducing social connections even further

```{r non.essl.net.3, cache=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
meandeg.non.essl.3 <- 0.8
fit.non.essl.3 <- ergm(emptynet~edges+concurrent, target.stats= c(meandeg.non.essl.2*n/2,0))
net.non.essl.3 <- simulate(fit.non.essl.3)
net.comb.3 <- net.non.essl.3
net.comb.3 <- add.edges(net.comb.3, tail=net.essl.el[,1], head=net.essl.el[,2])
plot(net.comb.3, vertex.cex=1.5, vertex.col=3+essl, edge.col="gray20")
```

# Steve's To Do list
2. Decide on authors
3. Homeless, blended households, relations are dynamic, non-discrete, etc.
