---
title: "**Can't I *please* just visit one friend?**"
subtitle: "Visualizing social distancing networks in the era of COVID-19"
author: 
 - Steven M. Goodreau, University of Washington
 - Emily D. Pollock, University of Washington
 - Jeanette K. Birnbaum, University of Washington
 - Deven T. Hamilton, University of Washington
 - Martina Morris, University of Washington
 - on behalf of the Statnet Development Team
date: "4/3/2020"
output:
  html_document:
    highlight: kate
    theme: lumen
    toc: yes
    toc_float: true
    number_sections: true
    fig_width: 7
    fig_height: 4
    fig_caption: true
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The situation

COVID-19 is raging around the world, and you and your family have been asked (OK, told) to stay home and practice social distancing.

For the most part you are doing so, because you are good people.

But, you wonder---or perhaps, more likely, your kids wonder---**is there really so much harm in meeting up with just one friend?** 

After all, there are all kinds of interactions still happening---between health care workers and patients, people in the food services industries and their customers, and so on.  Why does one more matter?

This is a question we've heard in a variety of forms, from our own friends and relatives.

**To be clear, when we talk about "meeting up" here, we mean doing so in a way that doesn't carefully follow all the rules about social distancing.** That is, you interact directly or at a distance of less than 6 feet, and/or you share various items between you without rigorous handwashing and disinfecting.

For many of the people asking the question, it seems particularly low-risk to meet up with a friend if neither one has an elderly person in their household, or someone else with a compromised immune system, given how concentrated the mortality is within those populations.

As network epidemiologists, this is our effort to explain why these connections matter more than they may seem.

<br>
<br>
<br>

# The Good Ol' Days {.tabset .tabset-pills .tabset-fade}

Let's start by imagining back to the good old days, just a few short weeks or months ago, before COVID-19 took off.  People had lots of daily interactions with lots of other people. They obviously varied in how many they had---some people had many, and some had few---but overall the number was a lot higher. Let's say, for the sake of the example, that the average household had 15 regular interactions with other households.

Now try to imagine what that might look like if you drew it. 

Often we visualize networks like this using diagrams with nodes and links.  Here we'll make one in which the **nodes (green dots) represent individual households**, and the **links (gray lines) represent frequent, repeated interactions** between any members of two households. 

We'll look at 200 households and 15 ties per household (the exact number doesn't matter much here---we just want to imagine a world that is much more connected than it is under social distancing).

What do we see?
 
[For those who want to read more detail---and including a few more terms and numbers---click on the "What's Going On"? button, here and in each example below. For those who just want to compare the pictures, continue on.]

## The Network

```{r pre-COVID, cache=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, fig.height=6, fig.width=6, fig.align='center'}
library(statnet)
set.seed(0)
n <- 200
emptynet <- network.initialize(n, directed=FALSE)
meandeg.precov <- 15
fit.precov <- ergm(emptynet~edges, target.stats=meandeg.precov*n/2)
net.precov <- simulate(fit.precov)
par(mai=c(0,0,0,0))
plot(net.precov, vertex.cex=1.5, mode="kamadakawai", vertex.col=3, edge.col="gray20")
largcomp.precov <- sum(component.largest(net.precov))
kpath.precov <- sna:::kpath.census(net.precov)$path.count[3,'Agg']/2
```

<br>
<br>
<br>
<br>

## What's going on? 

The connections among the nodes are so numerous that you can't even make sense of them---it's all just a single dense mass of ties. For a new virus that is able to spread across those ties, and to which everyone is susceptible, life is pretty good.

Let's put a few numbers on it for comparison to later scenarios:

- One important feature of a network that helps us to understand how far something might spread on it is the size of the **"largest cluster"**. That refers to the largest set of households who are all **reachable** to one another along any possible path---that is, there is at least one path of any length that connects the two households. 

- In this network, every household can be reached from every other household along at least one path---indeed, along many different ones---so the **largest cluster includes all 200 households---that is, 100% of households**.

- Another way of looking at connectivity is to count how many **paths** of different lengths there are.  For instance, if household A is connected to household B, which is in turn connected to household C, which is in turn connected to household D, then that would represent a 3-path, because there are three links (A-B-C-D).

- One can count up all the 2-paths in a network, or the 3-paths, or 4-paths, and so on. Each one gives a sense of how well connected the network is---how many different routes are there that infection can spread on.  For simplicity, let's focus on 3-paths. This is a distance that is long enough that you might imagine the person at the other end of it---for example, the daughter of the grocery store worker where your friend's parents shop, or your sister's friend's grandfather's doctor. If you became infected, you could imagine a chain of transmission that extends all the way to this person, even if you never actually find out that that person got infected. And with COVID-19, that could all happen before you have any symptoms and know that you're infected! 

- This network has **`r format(kpath.precov, scientific=F, big.mark=',')` different 3-paths**---that's a lot of routes for a virus to transmit three steps on!

[Note: The technical term in network theory for a group of connected people is a "component", but we think the idea of a "cluster" is more familar to most people]. 
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

# Perfect isolation {.tabset .tabset-pills .tabset-fade}

Now let's imagine a world at the other extreme---complete and utter lock-down, where somehow we manage to keep every person in their household and never interact with anyone else. What does that look like?

## The Network

```{r emptynet, cache=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align="center"}
par(mai=c(0,0,0,0))
plot(emptynet, vertex.cex=1.5, mode="kamadakawai", vertex.col=3, edge.col="gray20")
```

<br>
<br>
<br>
<br>

## What's going on? 
Here, the **largest cluster is of size 1**: every household is isolated from every other household. And of course there are **no 3-paths**.

What does that mean? Here the virus would not be able to spread from household to household at all, because there are no links. **No household is reachable from any other.** Anyone who had it might pass it onto the other members of their household, and each of those people would either die or recover.  But the virus could not spread to other households; it would quickly disappear.

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

# Adding in the essential workers {.tabset .tabset-pills .tabset-fade}

Of course, that world is impossible.  Why? Because some people work in jobs that are essential to maintain the safety, health and well-being of all of us: health care workers, first responders, grocery store workers, etc.

So, let's imagine what this might do to our network. 
Let's say that 1 in every 10 households contains someone with a job where they need to continue some connections. And the average number of connections they have is 4. 
If we **color the households with an essential worker as blue**, that gives us...

## The Network

```{r essl.net, cache=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align="center"}
meandeg.essl <- 4
prop.essl <- 0.1
essl <- rep(0, n)
essl[sample(1:n, round(prop.essl*n,0), replace=FALSE)] <- 1
set.vertex.attribute(emptynet, 'essl', essl)
fit.essl <- ergm(emptynet~edges+ nodematch("essl", diff=TRUE, levels=1),
              target.stats= c(
                meandeg.essl*prop.essl*n, 0),
              control=control.ergm(MCMC.burnin = 1e6))
net.essl <- simulate(fit.essl, control = control.simulate(MCMC.burnin = 1e6))
net.essl.el <- as.edgelist(net.essl)
par(mai=c(0,0,0,0))
plot(net.essl, vertex.cex=1.5, vertex.col=3+essl, edge.col="gray20")
largcomp.essl <- sum(component.largest(net.essl))
kpath.essl <- sna:::kpath.census(net.essl)$path.count[3,'Agg']/2

```
<br>
<br>
<br>
<br>

## What's going on? 
Even when States order people to "Stay Home", there are execptions for essential workers:

- We need doctors and nurses and EMTs interacting with people.
- We need all the people that get food to us---growing it, transporting it, selling it---to interact with people.
- We need the people who manage and maintain our water supply and electricity and gas to keep showing up at work.
- We need first responders keeping us safe.
- In big cities, we need transport workers helping all of these people get back and forth between work and home.
- And so on. 

It's hard to put real numbers on this, but we'll start with what seems like a reasonable example. Let's say they use hand washing and some distancing throughout the course of their work day as much as they can, so that they are able to bring their average effective contacts, in terms of reasonable transmission risk, down to 4.

You can see that the virus has some room to move around again---not nearly as much as in the days before social distancing, but still some. And it isn't just between the households with essential workers and their direct contacts.  Sometimes the same household will be connected to two essential workers, creating a longer path. Sometimes two households with essential workers will end up directly connected, putting their two sets of contacts into the same path.

Altogether, we see here a **largest cluster of size `r largcomp.essl`, or `r round(100*largcomp.essl/n, 1)`%, of households.** (Can you find it?) This is a lot smaller than 100%, and a fair bit bigger than 0%.

But households within the same cluster generally don't have lots of different paths by which they're connected.  So in total, there are **only `r kpath.essl` 3-paths.**  That's more than 0, but *much much much* less than 348,744!

Let us be very clear here: some people are going to get infected, and some people are going to die, because of these connections.  It's that simple.  

But these connections are so essential that we as a society are willing to make that trade-off. Without them, many more people would die of other basic things besides COVID-19: starvation, freezing, crime, other diseases.

And hopefully, the deaths here won't be that high. First off, most households are still effectively isolated. And second, those that are connected are only held together by a pretty sparse set of ties, as reflected by the low number of 3-paths. Since transmission is not 100% guaranteed, not all of those in the same cluster as an infected person will become infected. This set of households are all **reachable** from one another, but it's not guaranteed that they will all actually be reached by a virus present in that cluster. The fewer the different routes connecting an infected person and a susceptible person, the less likely transmission will reach from one to the other.

<br>
<br>
<br>
<br>
<br>
<br>

# Visiting just one friend {.tabset .tabset-pills .tabset-fade}

But, we all know now: strict social distancing starts to get boring after a while. And meeting up to hang out with just one person outside your household---a friend especially---is pretty tempting.  And it just doesn't seem like such a big deal when these other connections are already happening, right?

What happens if an average of two people in each household each decide to maintain an in-person social connection with one person from another household? 

## The Network 

```{r non.essl.net.1, cache=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align="center"}
meandeg.non.essl.1 <- 2
fit.non.essl.1 <- ergm(emptynet~edges, target.stats= meandeg.non.essl.1*n/2)
net.non.essl.1 <- simulate(fit.non.essl.1)
net.comb.1 <- net.non.essl.1
net.comb.1 <- add.edges(net.comb.1, tail=net.essl.el[,1], head=net.essl.el[,2])
par(mai=c(0,0,0,0))
plot(net.comb.1, vertex.cex=1.5, vertex.col=3+essl, edge.col="gray20")
largcomp.comb.1 <- sum(component.largest(net.comb.1))
kpath.comb.1 <- sna:::kpath.census(net.comb.1)$path.count[3,'Agg']/2
```

<br>
<br>
<br>
<br>
<br>

## What's going on?
Wow! This network is a lot more connected than the one with just essential workers.  

Most households have more than one person in them. In fact, the **average household size in the US is 2.6 people.**   Some households will have none, and some will have 3, but on average it's 2.6.

Let's say, then, that you decide to meet up with your friend Sue.  But if Sue's parents are letting her go out, then they probably also need to let her brother Ed go out to meet up with his friend.  Note that this is still less than the average household size---not everybody is meeting up with a friend.

What happens when we lay these new connections on top of the essential worker network?  A lot more connectivity.

Indeed, the **largest cluster contains `r largcomp.comb.1`, or `r round(100*largcomp.comb.1/n, 1)`%, of households.** That is **`r round(largcomp.comb.1/largcomp.essl, 1)` times larger than the essential network.**

And **the number of 3-paths here is `r kpath.comb.1`**.  That's **`r round(kpath.comb.1/kpath.essl, 1)` times larger than in the essential worker network!**  That's a lot more connections.  And therefore a lot more transmission, and a lot more death. Just so people can hang out with one friend!

If you are part of that large cluster, then now your household can reach `r round(100*largcomp.comb.1/n, 1)`% of the households in your community, and they can reach you! You may not even know the people in these households, and you won't even know whether you're part of that group or not. But the virus can easily travel along the network of connections to get from them to you, especially when there are multiple paths through the network. It's a very connected, and effective, network of transmission.

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

# Visiting just one friend per household {.tabset .tabset-pills .tabset-fade}

OK, maybe households with multiple people could all agree that only one person can hang out with one friend. This seems like it should have a lot less connectivity, right? 

## The Network

```{r non.essl.net.2, cache=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align="center"}
meandeg.non.essl.2 <- 0.999
fit.non.essl.2 <- ergm(emptynet~edges+ concurrent, target.stats= c(meandeg.non.essl.2*n/2, 0))
net.non.essl.2 <- simulate(fit.non.essl.2, control=control.simulate(MCMC.burnin=1e8))
net.comb.2 <- net.non.essl.2
net.comb.2 <- add.edges(net.comb.2, tail=net.essl.el[,1], head=net.essl.el[,2])
par(mai=c(0,0,0,0))
plot(net.comb.2, vertex.cex=1.5, vertex.col=3+essl, edge.col="gray20")
largcomp.comb.2 <- sum(component.largest(net.comb.2))
kpath.comb.2 <- sna:::kpath.census(net.comb.2)$path.count[3,'Agg']/2
```

<br>
<br>
<br>
<br>

## What's going on? 

Maybe parents don't feel like they need friend time as much as their kid does. Of course, if there are multiple kids, that creates a problem---eveyone will want to be the one who gets to hang out with their friend!  But let's say we could make it work.  

To examine this, then, we won't just set an *average* of one non-essential tie per household; we'll make every household have exactly one, and no more. Pairs of households become connected, but then they don't create the same chains that are possible when there are two connections per household. Bu this is still in addition to the essential worker connections. Does this create much less connectivity than the case of two friends per household?  Not really.

This still creates a fair amount of connection---the **largest cluster contains `r largcomp.comb.2`.** That's still the majority of households, at  **`r round(100*largcomp.comb.2/n, 1)`%,** and **`r round(largcomp.comb.2/largcomp.essl, 1)` times larger than the essential network.**

And there are still **`r kpath.comb.2` 3-paths**, or **`r round(kpath.comb.2/kpath.essl, 1)` times more than in the essential worker network alone.** That still allows for a lot of extra transmissions and illnesses and hospitalizations and deaths.

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

# Lessons learned

**Humans are social animals, and reducing connectivity in social networks is hard.** 
And some connections just can't broken at all if we want to maintain the basic functioning of the systems we need in place to all live. Other connections laid on top of those create more network connectivity than you might think. And thus:

**Every additional connection that we can postpone until COVID-19 is under control has the potential to save one or more lives.**  Yes, **every one**.  

You may never know whether it made a difference, and if it did, exactly whose life was saved. 

- They may be multiple degrees of separation away from you. 

- They may be your friend's mother's patient's father. 

- They may be your sister's friend's grandfather's doctor. 

- They may be the daughter of the grocery store worker where your friend's parents shop. 

**Regardless of who it is---waiting a bit to hang out with your friends again is worth it!**

<br>
<br>
<br>
<br>

# The fine print for those who are interested {.tabset}

## Model caveats

Now, you may be thinking -- isn't this picture a bit simplistic, in a bunch of ways? Not everybody fits neatly into households; for example, some people live in group quarters like dorms or nursing homes, some kids divide their time between two parental households, and some people are homeless. Surely there is a lot more variation in connections among people than you seem to have here. It seems that, beyond a few simple rules about numbers of connections per household, you are assuming that households pick each other randomly, and surely there is a lot more structure to real networks than that. And surely relationships are actually dynamic---they form and break over time, and not everyone sees the same people from the same households every day.  

All of those points are true. But we put them aside and purposefully simplified the world a bit so that we could gain some clear insight about those scenarios first. This is a common approach in scientific investigation in general, and network epidemiology in particular.  Once one gains those insights, then one can begin adding in more complexity. Actual models used in the field of network epidemiology typically have a lot more complexity in them than the versions we look at here---but these are great for communicating basic principles to a broad audience.

<br>
<br>
<br>
<br>

## Model code

For those of you who find this interesting and have some coding skills in R, we include all of the code for all of these examples for you to look at and play with and change values for or add complexity to, as you wish.  Note that these examples use the *statnet* suite of R packages.  Thank you goes to the full **statnet development team** for all of these. Full citation information is:

Mark S. Handcock, David R. Hunter, Carter T. Butts, Steven M. Goodreau, Pavel N. Krivitsky, Skye Bender-deMoll, and Martina Morris (2019). statnet: Software tools for the Statistical Modeling of Network Data. versin 2019.6. URL http://statnet.org, https://cran.r-project.org/web/packages/statnet/.



```{r full-code, eval=FALSE}

## setup
library(statnet)
n <- 200
emptynet <- network.initialize(n, directed=FALSE)

## pre-COVID network
meandeg.precov <- 15
fit.precov <- ergm(emptynet~edges, target.stats=meandeg.precov*n/2)
net.precov <- simulate(fit.precov)
plot(net.precov, vertex.cex=1.5, mode="kamadakawai", vertex.col=3,
     edge.col="gray20")
largcomp.precov <- sum(component.largest(net.precov))

## empty network
plot(emptynet, vertex.cex=1.5, mode="kamadakawai", vertex.col=3,
     edge.col="gray20")

## essential network
meandeg.essl <- 4
prop.essl <- 0.1
essl <- rep(0, n)
essl[sample(1:n, round(prop.essl*n,0), replace=FALSE)] <- 1
set.vertex.attribute(emptynet, 'essl', essl)
fit.essl <- ergm(emptynet~edges+ nodematch("essl", diff=TRUE, levels=1),
              target.stats= c(meandeg.essl*prop.essl*n, 0),
              control=control.ergm(MCMC.burnin = 1e6))
net.essl <- simulate(fit.essl, control = control.simulate(MCMC.burnin = 1e6))
net.essl.el <- as.edgelist(net.essl)
plot(net.essl, vertex.cex=1.5, vertex.col=3+essl, edge.col="gray20")
largcomp.essl <- sum(component.largest(net.essl))
kpath.essl <- sna:::kpath.census(net.essl)$path.count[3,'Agg']/2

## non.essl.net.1
meandeg.non.essl.1 <- 2
fit.non.essl.1 <- ergm(emptynet~edges, target.stats= meandeg.non.essl.1*n/2)
net.non.essl.1 <- simulate(fit.non.essl.1)
net.comb.1 <- net.non.essl.1
net.comb.1 <- add.edges(net.comb.1, tail=net.essl.el[,1], head=net.essl.el[,2])
plot(net.comb.1, vertex.cex=1.5, vertex.col=3+essl, edge.col="gray20")
largcomp.comb.1 <- sum(component.largest(net.comb.1))
kpath.comb.1 <- sna:::kpath.census(net.comb.1)$path.count[3,'Agg']/2

## non.essl.net.2
meandeg.non.essl.2 <- 1
fit.non.essl.2 <- ergm(emptynet~edges+ concurrent, target.stats= c(meandeg.non.essl.2*n/2, 0))
net.non.essl.2 <- simulate(fit.non.essl.2, control=control.simulate(MCMC.burnin=1e8))
net.comb.2 <- net.non.essl.2
net.comb.2 <- add.edges(net.comb.2, tail=net.essl.el[,1], head=net.essl.el[,2])
plot(net.comb.2, vertex.cex=1.5, vertex.col=3+essl, edge.col="gray20")
largcomp.comb.2 <- sum(component.largest(net.comb.2))
kpath.comb.2 <- sna:::kpath.census(net.comb.2)$path.count[3,'Agg']/2
```
